\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-08 honors_thesis/journal_paper/paper.tex
updated references in journal_paper

% AML bibliography references
\newcommand{\bib}[1]{/aml/home/mjg82/bib/#1}

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{mcnabb-cec07,belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
and asynchronous algorithms
\cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
20,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
effect on the value obtained.  However, increasing the swarm size from 100 to
200 makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  For this reason, previous work has recommended
the use of a swarm size of 50 for this function~\cite{bratton-sis07}.  Thus, in
at least some cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere.eps}
  \caption{Function Sphere with a swarm with various swarm sizes.}
  \label{fig:evals-sphere}
\end{figure}

@@

%\begin{quote}
%SKIP SKIP
%
%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
%algorithms.  When the next sampling position of the algorithm does not depend
%on the value of the objective function at the current position,
%future positions can be speculated and evaluated in parallel with the current
%position.
%This is trivially true for optimization by random sampling, for example.
%The next random sample does not depend on the previous point nor on the value
%of the objective function at that point in any way.
%It depends only on a random number.
%There is no need to wait to see
%what objective function value was obtained at the previous point,
%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
%In other optimization algorithms, the next position depends on the previous position, but not the
%value obtained there, a random walk for example.
%In this case
%the next position to be evaluated
%depends only on the previous position and a random number.
%In this case, as it was in the case of random sampling,
%the object function evaluation of the next position can be done in parallel
%because it 
%does not depend on the \emph{value} of the objective function at the previous position.
%There are an infinite number of possible next positions, but only one given the draw of the random number.
%
%In still other cases, the next position to be evaluated may depend on the current position and some
%discrete valued (binary, for example) test on the value obtained at the current position.
%For example, consider a greedy random walk, in which the state only advances if a better
%position is found.
%That is, if the current value is better than the previous value,
%use the sum of the current position and the random number as the next position.
%Otherwise, computer the next position using the sum of the previous position and the random number.
%In this case the objective function can be evaluated at three positions:
%(1) the value of the objective function at the current position,
%(2) the value at the position if the test turns out to be true and (3) the value at the position
%if the test turns out to be false; all in parallel.
%In this example it is important to note that we do not need the value of the objective function at the current
%point to to reduce the infinite number of possible next positions to just two alternatives.
%Choosing between these two alternatives can be done post-hoc, after the objective function value
%at the current position and at the two alternative are computed.
%For all but the most computationally-trivial objective functions, this approach saves almost half of the
%``wall-clock'' time.
%This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
%depends on positions and a binary test.
%
%This approach can not be used in gradient methods where the
%step size for sampling typically depends on the gradient and therefor the position and 
%a set of infinitely many possible \emph{values} outcomes
%obtained at the current position (not binary or even finite).
%\end{quote}

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently approximate the optimal solution
to a given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies a particle is also a
neighbor to itself in that its own position and value are considered when
updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
using the Ring topology each particle with index $i$ has three neighbors:
$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
$p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less communication~\cite{mcnabb-cec09}.  In the
Random topology, each particle randomly picks two other particles to share
information with at each iteration, along with itself.  Thus in both the Ring
and the Random topologies, all particles have three neighbors.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In~\sec{results} we report the results for performing
speculative evaluation using both the Ring topology and the Random topology on
a number of common benchmark functions.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
objective function at all necessary positions, the master then decides which
position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
The outline of the procedure is given in \alg{centralized}.

@@

Lastly we look at a function that does not show any kind of improvement with
this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
swarms get caught in local optima.  Because our speculative algorithm has a
significantly smaller swarm size, it gets stuck at higher values while the
larger swarms performing regular PSO continue to improve the best value found.
\fig{rastrigin} shows the results graphically.

@@

We gave a passing glance to some very interesting statistical characteristics
of PSO.  Our motivation was to use the statistics to improve our speculative
algorithm, but it seems there are some nuggets in the numbers that could be
used to improve our understanding of the behavior of PSO in general, and what
role a topology plays.  A potentially very fruitful avenue of future work would
be to further analyze the branch statistics in PSO across many functions and
topologies, categorizing both the topologies and the functions and discovering
what it is that makes certain topologies work well with certain functions.
There are also some particles that seem to drive the swarm and others that are
stagnant almost the entire time.  Such information could be used in a
TRIBES-like fashion~\cite{clerc-oep03} to determine which particles should be
removed from the swarm, perhaps starting them close to well-performing
particles.

@@

A technique similar to ours has been previously used to parallelize simulated
annealing~\cite{witte-tpds91}.  Any other optimization algorithm that only
depends on current sampling positions when computing the next position to
sample can be parallelized with this technique.  In particular, genetic
algorithms produce future generations by combining individuals from the current
generation.  With a large population size there would be an unwieldy amount of
possible future individuals, but the potential exists to modify the algorithm
to use some kind of speculative evaluation.  A related but quite distinct
technique for reducing computation in evolutionary algorithms was proposed
in~\cite{poli-ai06}.

@@

\bibliographystyle{unsrt}
\bibliography{%
\bib{wolpert-tec97},%
\bib{merris-01},%
\bib{pso/clerc-tec02},%
\bib{pso/kennedy-icnn95},%
\bib{pso/bratton-sis07},%
\bib{pso/largeswarms/perez-emc05},%
\bib{pso/largeswarms/montes-gecco08.bib},%
\bib{pso/topology/liang-sis05.bib},%
\bib{pso/parallel/belal-ijicis04},%
\bib{pso/parallel/chu-sci06},%
\bib{pso/parallel/jin-aps05},%
\bib{pso/parallel/koh-ijnme06},%
\bib{pso/parallel/mostaghim-report06},%
\bib{pso/parallel/parsopoulos-aia04},%
\bib{pso/parallel/schutte-ijnme04},%
\bib{pso/parallel/venter-wcsmo05},%
\bib{ga/poli-ai06},%
\bib{sim-anneal/parallel/witte-tpds91.bib},%
\bib{pso/topology/kennedy-cec02.bib},%
\bib{pso/topology/mohais-cec04.bib},%
\bib{pso/topology/jordan-gecco08.bib},%
\bib{pso/topology/clerc-oep03.bib},%
\bib{pso/topology/miranda-ijcir08.bib},%
\bib{pso/topology/mendes-phd04.bib},%
\bib{aml/mcnabb-cec07},%
\bib{aml/mcnabb-cec09}}

>>>>

@@

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{mcnabb-2007-parallel-pso-using-mapreduce,belal-2004-parallel-models-for-pso,chu-2006-intelligent-parallel-pso,jin-2005-pso-antenna-designs,parsopoulos-2004-parallel-vector-evaluated-pso,schutte-2004-parallel-global-optimization-with-pso}
and asynchronous algorithms
\cite{koh-2006-parallel-asynchronous-pso,mostaghim-2006-multi-objective-pso-on-grids,venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

For many functions there comes a point of diminishing returns with respect to
adding particles.  In \fig{evals-sphere} we show the value obtained after
20,000 function evaluations (not iterations) as a function of swarm size for
the well-know benchmark function Sphere (20 dimensions, reporting the average
of thirty runs).  Increasing the swarm size from 10 to 50 has a significant
effect on the value obtained.  However, increasing the swarm size from 100 to
200 makes the algorithm less efficient, that is, it reduces the progress the
algorithm makes per evaluation.  For this reason, previous work has recommended
the use of a swarm size of 50 for this function~\cite{bratton-2007-defining-a-standard-for-pso}.  Thus, in
at least some cases, adding particles indefinitely will not yield an efficient
implementation.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with a swarm with various swarm sizes.}
  \label{fig:evals-sphere}
\end{figure}

@@

%\begin{quote}
%SKIP SKIP
%
%No really, try reading without this! The idea of speculative execution can be applied to some types of optimization
%algorithms.  When the next sampling position of the algorithm does not depend
%on the value of the objective function at the current position,
%future positions can be speculated and evaluated in parallel with the current
%position.
%This is trivially true for optimization by random sampling, for example.
%The next random sample does not depend on the previous point nor on the value
%of the objective function at that point in any way.
%It depends only on a random number.
%There is no need to wait to see
%what objective function value was obtained at the previous point,
%evaluation of the object function at the next point can begin as soon as the next random number is drawn.
%In other optimization algorithms, the next position depends on the previous position, but not the
%value obtained there, a random walk for example.
%In this case
%the next position to be evaluated
%depends only on the previous position and a random number.
%In this case, as it was in the case of random sampling,
%the object function evaluation of the next position can be done in parallel
%because it 
%does not depend on the \emph{value} of the objective function at the previous position.
%There are an infinite number of possible next positions, but only one given the draw of the random number.
%
%In still other cases, the next position to be evaluated may depend on the current position and some
%discrete valued (binary, for example) test on the value obtained at the current position.
%For example, consider a greedy random walk, in which the state only advances if a better
%position is found.
%That is, if the current value is better than the previous value,
%use the sum of the current position and the random number as the next position.
%Otherwise, computer the next position using the sum of the previous position and the random number.
%In this case the objective function can be evaluated at three positions:
%(1) the value of the objective function at the current position,
%(2) the value at the position if the test turns out to be true and (3) the value at the position
%if the test turns out to be false; all in parallel.
%In this example it is important to note that we do not need the value of the objective function at the current
%point to to reduce the infinite number of possible next positions to just two alternatives.
%Choosing between these two alternatives can be done post-hoc, after the objective function value
%at the current position and at the two alternative are computed.
%For all but the most computationally-trivial objective functions, this approach saves almost half of the
%``wall-clock'' time.
%This approach has been applied to simulated annealing~\cite{witte-1991-parallel-simulated-annealing-speculative} which like the greedy random walk,
%depends on positions and a binary test.
%
%This approach can not be used in gradient methods where the
%step size for sampling typically depends on the gradient and therefor the position and 
%a set of infinitely many possible \emph{values} outcomes
%obtained at the current position (not binary or even finite).
%\end{quote}

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-1995-particle-swarm-optimization}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently approximate the optimal solution
to a given objective function.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-2007-defining-a-standard-for-pso}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-2002-constricted-pso}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies a particle is also a
neighbor to itself in that its own position and value are considered when
updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
using the Ring topology each particle with index $i$ has three neighbors:
$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
$p$ neighbors.

In this paper we use these topologies as well as a parallel adaptation of the
Complete topology, called Random, that has been shown to approximate the
behavior of Complete with far less communication~\cite{mcnabb-2009-large-particle-swarms}.  In the
Random topology, each particle randomly picks two other particles to share
information with at each iteration, along with itself.  Thus in both the Ring
and the Random topologies, all particles have three neighbors.

@@

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-2009-large-particle-swarms}, which has been shown to effectively simulate the
Complete topology.  In~\sec{results} we report the results for performing
speculative evaluation using both the Ring topology and the Random topology on
a number of common benchmark functions.

@@

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform speculative evaluation in such
an architecture, the master generates the positions to evaluate speculatively
as in Equation~\eqref{eq:pos2update}.  After having the slaves evaluate the
objective function at all necessary positions, the master then decides which
position to accept for each particle, as in Equation~\eqref{eq:val2updatelong}.
The outline of the procedure is given in \alg{centralized}.

@@

Lastly we look at a function that does not show any kind of improvement with
this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller
swarms get caught in local optima.  Because our speculative algorithm has a
significantly smaller swarm size, it gets stuck at higher values while the
larger swarms performing regular PSO continue to improve the best value found.
\fig{rastrigin} shows the results graphically.

@@

We gave a passing glance to some very interesting statistical characteristics
of PSO.  Our motivation was to use the statistics to improve our speculative
algorithm, but it seems there are some nuggets in the numbers that could be
used to improve our understanding of the behavior of PSO in general, and what
role a topology plays.  A potentially very fruitful avenue of future work would
be to further analyze the branch statistics in PSO across many functions and
topologies, categorizing both the topologies and the functions and discovering
what it is that makes certain topologies work well with certain functions.
There are also some particles that seem to drive the swarm and others that are
stagnant almost the entire time.  Such information could be used in a
TRIBES-like fashion~\cite{clerc-2003-tribes} to determine which particles should be
removed from the swarm, perhaps starting them close to well-performing
particles.

@@

A technique similar to ours has been previously used to parallelize simulated
annealing~\cite{witte-1991-parallel-simulated-annealing-speculative}.  Any other optimization algorithm that only
depends on current sampling positions when computing the next position to
sample can be parallelized with this technique.  In particular, genetic
algorithms produce future generations by combining individuals from the current
generation.  With a large population size there would be an unwieldy amount of
possible future individuals, but the potential exists to modify the algorithm
to use some kind of speculative evaluation.  A related but quite distinct
technique for reducing computation in evolutionary algorithms was proposed
in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.

@@

\bibliographystyle{unsrt}
\bibliography{%
../../../bib/aml/bib,%
../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/sim-anneal/parallel/bib,%
../../../bib/ga/bib,%
}
